---
title: "Introduction to Statistical Learning"
subtitle: "Resampling Methods: The Bootstrap"
author: "Course Instructor"
institute: "Department of Mathematics"
date: today
format: 
  beamer:
    theme: Madrid
    colortheme: default
    keep-tex: true
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{positioning, calc}
execute:
  warning: false
  message: false
---

## What is the Bootstrap?

* The **bootstrap** is a widely applicable and extremely powerful statistical tool.
* It is primarily used to quantify the uncertainty associated with a given estimator.
* **Common use cases:**
  * Estimating the standard errors of an estimator.
  * Constructing confidence intervals of a parameter.
  * Evaluating the performance of a complex machine learning model.
* **Core Idea:** To simulate drawing new samples from a population by resampling using original data.

## Motivating Example: Asset Allocation

* Suppose we wish to invest a fixed sum of money in two financial assets that yield returns $X$ and $Y$.
* We will invest a fraction $\alpha$ in $X$ and $(1-\alpha)$ in $Y$.
* **Goal:** Choose $\alpha$ to minimize the total risk, or variance, of our investment.
* The variance of the return is:
$$
\text{Var}(\alpha X + (1-\alpha)Y) = \alpha^2 \sigma_X^2 + (1-\alpha)^2 \sigma_Y^2 + 2\alpha(1-\alpha)\sigma_{XY}
$$
* The value that minimizes this risk is:
$$
\alpha = \frac{\sigma_Y^2 - \sigma_{XY}}{\sigma_X^2 + \sigma_Y^2 - 2\sigma_{XY}}
$$

## The Challenge in Practice

* $\sigma_X^2, \sigma_Y^2$, and $\sigma_{XY}$ are generally **unknown**.
* However, we can compute estimates ($\hat{\sigma}_X^2, \hat{\sigma}_Y^2, \hat{\sigma}_{XY}$) using a sample of past returns.
* We then estimate the optimal allocation:
$$
\hat{\alpha} = \frac{\hat{\sigma}_Y^2 - \hat{\sigma}_{XY}}{\hat{\sigma}_X^2 + \hat{\sigma}_Y^2 - 2\hat{\sigma}_{XY}}
$$
* **Question:** How accurate is $\hat{\alpha}$ as an estimate of $\alpha$?
  <!-- What is its standard error? -->
* The bootstrap provides a simple way to answer.

## The Bootstrap Procedure

Instead of obtaining independent data sets from the true population (which is usually impossible), we obtain distinct data sets by repeatedly **sampling observations from the original data set with replacement**.

**The Algorithm:**

1. Treat your original sample of size $n$ as the "population."
2. Draw a new sample of size $n$ from this original sample *with replacement*. This is a **bootstrap data set** ($Z^{*1}$).
3. Calculate the statistic of interest (e.g., $\hat{\alpha}^{*1}$) using this bootstrap data set.
4. Repeat steps 2 and 3 a large number of times ($B$ times, e.g., $B = 1000$).
5. You now have $B$ different bootstrap estimates: $\hat{\alpha}^{*1}, \hat{\alpha}^{*2}, \dots, \hat{\alpha}^{*B}$.

## Schematic of the Bootstrap Process

\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    node distance=1.2cm and 0.5cm,
    box/.style={rectangle, draw, rounded corners, align=center, fill=blue!10, minimum width=2.5cm, minimum height=1cm},
    estimate/.style={rectangle, draw, rounded corners, align=center, fill=orange!10, minimum width=2.5cm, minimum height=0.8cm},
    arrow/.style={->, thick, >=stealth}
]

% Original Data
\node[box, fill=green!10] (orig) {\textbf{Original Data} \\ $Z = (X_1, \dots, X_n)$};

% Bootstrap Samples
\node[box] (boot2) [below=of orig] {Bootstrap Sample 2 \\ $Z^{*2}$};
\node[box] (boot1) [left=of boot2, xshift=-0.5cm] {Bootstrap Sample 1 \\ $Z^{*1}$};
\node (dots) [right=of boot2, xshift=0.5cm] {\Large $\dots$};
\node[box] (bootB) [right=of dots, xshift=0.5cm] {Bootstrap Sample $B$ \\ $Z^{*B}$};

% Arrows from Original to Samples
\draw[arrow] (orig) -- (boot1.north);
\draw[arrow] (orig) -- (boot2.north);
\draw[arrow] (orig) -- (bootB.north);

% Estimates
\node[estimate] (est1) [below=of boot1] {Statistic $\hat{\alpha}^{*1}$};
\node[estimate] (est2) [below=of boot2] {Statistic $\hat{\alpha}^{*2}$};
\node (dots2) [below=of dots] {\Large $\dots$};
\node[estimate] (estB) [below=of bootB] {Statistic $\hat{\alpha}^{*B}$};

% Arrows from Samples to Estimates
\draw[arrow] (boot1) -- (est1);
\draw[arrow] (boot2) -- (est2);
\draw[arrow] (bootB) -- (estB);

% Calculation of center coordinate for correct alignment
\coordinate (row_center) at ($(est1.west)!0.5!(estB.east)$);

% Final Distribution
\node[box, fill=red!10, minimum width=14cm] (dist) at ([yshift=-2.5cm]row_center) {\textbf{Bootstrap Distribution} \\ Use to compute Standard Error $\text{SE}_B(\hat{\alpha})$, Confidence Intervals, etc.};

% Arrows from Estimates to Final Distribution
\draw[arrow] (est1) -- (dist.north -| est1.south);
\draw[arrow] (est2) -- (dist.north -| est2.south);
\draw[arrow] (estB) -- (dist.north -| estB.south);

\end{tikzpicture}
}
\end{center}

## Estimating the Standard Error

* Once we have our $B$ bootstrap estimates, we can analyze their distribution.
* The **bootstrap standard error** is simply the sample standard deviation of the bootstrap estimates:
$$
\text{SE}_B(\hat{\alpha}) = \sqrt{\frac{1}{B-1} \sum_{i=1}^B \left( \hat{\alpha}^{*i} - \frac{1}{B} \sum_{i'=1}^B \hat{\alpha}^{*i'} \right)^2}
$$
* This serves as an estimate of the standard error of $\hat{\alpha}$ estimated from the original data set.

## Visualizing the Bootstrap Distribution

* After generating $B = 1000$ bootstrap estimates of $\hat{\alpha}$, we can plot their overall shape.
* Notice that the distribution looks approximately normal, centered around our original estimate.

```{r}
#| label: boot-hist
#| echo: false
#| fig-align: center
#| fig-width: 6
#| fig-height: 3
library(ISLR2)
library(boot)

# 1. Define the function to calculate alpha
alpha.fn <- function(data, index) {
  X <- data$X[index]
  Y <- data$Y[index]
  return((var(Y) - cov(X, Y)) / 
         (var(X) + var(Y) - 2 * cov(X, Y)))
}

# 2. Run the bootstrap with B = 1000
set.seed(1)
boot.out <- boot(Portfolio, alpha.fn, R = 1000)
hist(boot.out$t, breaks = 30, col = "lightblue", 
     main = "",
     xlab = "Bootstrap Estimates", 
     border = "white", freq = FALSE)
curve(dnorm(x, mean(boot.out$t), sd(boot.out$t)), 
      add = TRUE, col = "red", lwd = 2)
```

## Visualizing the Spread (Boxplot)

* A boxplot provides a concise summary of the same distribution, highlighting the median, quartiles, and any potential outliers.
* The interquartile range is closely related to $\text{SE}_B(\hat{\alpha})$.

```{r}
#| label: boot-box
#| echo: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

boxplot(boot.out$t, col = "orange", 
        main = "",
        ylab = "Bootstrap Estimates")
```

## Two Flavors of the Bootstrap

**1. Non-Parametric Bootstrap**

* Makes **no assumptions** about the underlying population distribution.
* Resamples directly from the observed data (the empirical distribution).
* **Use when:** The true distribution is unknown, complex, or you cannot confidently commit to a parametric model.

**2. Parametric Bootstrap**

* Assumes the data comes from a **known parametric family** of distributions (e.g., Normal, Exponential, Poisson).
* Estimates the parameters from the data, then draws new samples from the fitted theoretical distribution.
* **Use when:** You have strong theoretical or empirical reasons to assume a specific probability distribution.

## Non-Parametric Bootstrap: Example

**Goal:** Estimate the standard error of the sample median.

**Observed Data:** A sample of $n=5$: $X = \{3, 8, 2, 10, 5\}$.  
**Sample Median:** $5$

**The Procedure:**

1. **Resample with replacement** directly from $X$ to create $B$ bootstrap datasets of size $n=5$.
   * Sample $X^{*1} = \{3, 3, 10, 2, 8\} \implies \text{Median}^{*1} = 3$
   * Sample $X^{*2} = \{8, 8, 5, 2, 5\} \implies \text{Median}^{*2} = 5$
   * Sample $X^{*3} = \{10, 10, 10, 3, 2\} \implies \text{Median}^{*3} = 10$
2. **Calculate:** Compute the median for all $B$ (e.g., $B=1000$) samples.
3. **Evaluate:** Find the standard deviation of these $1000$ medians to get the standard error, $\text{SE}_B(\text{Median})$.

## Parametric Bootstrap: Example

**Goal:** Estimate the standard error of the mean time to failure.

**Observed Data:** Lifetimes of $n=10$ components. We **assume** the lifetimes follow an Exponential distribution: $X \sim \text{Exp}(\lambda)$.

**The Procedure:**

1. **Estimate Parameter:** Calculate the maximum likelihood estimate for the rate parameter from the original data, say $\hat{\lambda} = 0.05$.
2. **Simulate from Model:** Generate $B$ new datasets of size $n=10$ directly from $\text{Exp}(\lambda = 0.05)$ using a random number generator.
   * Sample $X^{*1} \sim \text{Exp}(0.05) \implies \bar{X}^{*1} = 22.4$
   * Sample $X^{*2} \sim \text{Exp}(0.05) \implies \bar{X}^{*2} = 18.1$
3. **Calculate & Evaluate:** Compute the sample mean for each simulated dataset, then find the standard deviation to get $\text{SE}_B(\bar{X})$.

## Summary of the Bootstrap

* The bootstrap allows us to compute standard errors and confidence intervals for estimators where mathematical derivations are impossible or complex.
* It relies on the assumption that the original sample is a good representation of the underlying population.
* **Key Takeaway:** We mimic the process of obtaining new data sets from the population by obtaining new data sets from our original sample.
